The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2016-08-23 script/en/chapters/primes.tex
CT Book ed 12: Major update -- many changes, corrections, and additions during last 4 months

% $Id$
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% C H A P T E R   T H R E E :  P R I M E S
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage
% \hypertarget{Kapitel_2}{}
\chapter{Prime Numbers}
\label{Label_Kapitel_Primes}
(Bernhard Esslinger, May 1999; Updates Nov. 2000, Dec. 2001, June 2003, May 2005, March 2006, June 2007, Jan. 2010, Aug. 2013, Mar. 2016)

@@

\noindent Prime numbers are particularly important for three reasons:
\begin{itemize}
  \item In number theory, they are considered to be the basic components of
natural numbers, upon which numerous brilliant mathematical ideas are based.
  \item They are of extreme practical importance in modern
\index{cryptography!modern} cryptography (public key \index{cryptography!public key} cryptography). The most common public key procedure, invented at the end of
the 1970's, is \index{RSA} RSA encryption. Only using (large) prime numbers for
particular parameters can you guarantee that an algorithm is secure, both for
the RSA procedure and for even more modern procedures (e.g. elliptic curves).
  \item The search for the largest known prime numbers does not have any
practical usage known to date, but requires the best computers, is an excellent
benchmark (possibility for determining the performance\index{performance} of computers) and leads
to new calculation methods on many computers \\
(see also: \url{http://www.mersenne.org/prime.htm}).
\end{itemize}
Many people have been fascinated by prime numbers over the past two millennia.
Ambition to make new discoveries about prime numbers has often resulted in
brilliant ideas and conclusions. The following section provides an easily
comprehensible introduction to the basics of prime numbers. We will also
explain what is known about the distribution (density, number of prime numbers
in particular intervals) of prime numbers and how prime number tests work.

@@

\begin{figure}[!hb]
\begin{center}
\includegraphics[scale=0.3]{figures/Numberrectangle-with-colored-primes_Irisprime.jpg}
\caption[Primes between the first 390 integers -- marked with color]
        {Primes between the first 390 integers -- marked with color\protect\footnotemark}
\label{Primes-in-a-390-integer-rechtangle-figure}
\end{center}
\end{figure}
\footnotetext{%
   Graphics from 
   \url{http://mathforum.org/mathimages/index.php/Image:Irisprime.jpg}, 30*13 rectangle
   }

\begin{figure}[!hb]
\begin{center}
\includegraphics[scale=0.41]{figures/Ulam_spiral_via-CT2-within-1-999.png}
\caption[Primes between the first 999 integers -- as Ulam spiral]
        {Primes between the first 999 integers -- as Ulam spiral\protect\footnotemark}
\label{Primes-in-a-999-integer-ulam-spiral-figure}
\end{center}
\end{figure}
\footnotetext{%
   Graphics from CT2, menu Crypto Tutorials, Distribution of primes, Ulam's spiral; 32*32 points
   }
%\newpage
\begin{figure}[!hb]
\begin{center}
\includegraphics{figures/Ulam_spiral_Wikipedia.png}
\caption[Primes between the first 4000 integers -- as Ulam spiral]
        {Primes between the first 4000 integers -- as Ulam spiral\protect\footnotemark}
\label{Primes-in-a-4000-integer-ulam-spiral-figure}
\end{center}
\end{figure}
\footnotetext{%
   Graphics from 
   \url{http://mathforum.org/mathimages/index.php/Image:Ulam_spiral.png}, 200*200 Ulam spiral
   }

@@

% --------------------------------------------------------------------------
\section{How many prime numbers are there?}

For the natural numbers, the primes can be compared to elements in chemistry or
the elementary particles in physics (see \cite[p. 22]{pr:Blum1999}).

Although there are only $92$ natural chemical elements, the number of prime
numbers is unlimited. Even the Greek, \index{Euclid} Euclid%
\footnote{Euclid,
a Greek mathematician of 4th and 3rd century B.C. He worked at the
Egyptian academy of Alexandria and wrote ``The Elements'', the most well 
known systematically textbook of the Greek mathematics.} 
knew this in the third century B.C.
\begin{theorem}[Euclid\footnote{The common usage of the term does not denote Euclid as the inventor of the theorem rather;
the true inventor is merely not as prominent. The theorem has already been distinguished
and proven in Euclid's Elements (Book IX, theorem 20). The phraseology is remarkable due to 
the fact that the word infinite is not used. The text reads as followed
$$
O\acute{\iota}~\pi\varrho\tilde{\omega}\tau o \iota~\grave{\alpha}\varrho\iota\vartheta\mu o\grave{\iota}~
\pi\lambda\varepsilon\acute{\iota}o \upsilon\varsigma~\varepsilon\grave{\iota}\sigma\grave{\iota}~
\pi\alpha\nu\tau\grave{o}\varsigma~\tau o \tilde{\upsilon}~
\pi\varrho o \tau\varepsilon\vartheta\acute{\varepsilon}\nu\tau o \varsigma~
\pi\lambda\acute{\eta}\vartheta\ o \upsilon\varsigma~
\pi\varrho\acute{\omega}\tau\omega\nu~
\grave{\alpha}\varrho\iota\vartheta\mu\tilde{\omega}\nu,
$$
the English translation of which is: the prime numbers are more than 
any previously existing amount of prime numbers.
}]\label{thm-pz-euklid} % Ende der Fuﬂnote
% Fussnote VOR "]" in [Euclid], damit kein Leerraum vor der Fussnotennummer.
% Vorher stand da: ...(Euclid). BLANK Fussnote und das Blank stoerte.
% Nun steht die Fussnote direkt hinter "Euclid" und vor der ")".
% Eigentlich h‰tte ich sie gerne direkt hinter "(Euclid)", noch vor dem 
% automatisch gesetzten Punkt. 
The sequence of prime numbers does not discontinue.
Therefore, the quantity of prime numbers is infinite.
\end{theorem}
His proof that there is an infinite number of
primes is still considered to be a brilliant mathematical consideration and
conclusion today (\textbf{proof by contradiction} \index{proof by contradiction}). 
He assumed that there is only a finite number of primes and therefore a 
largest prime number. Based on this assumption, he drew logical conclusions
until he obtained an obvious contradiction. This meant that something must
be wrong. As there were no mistakes in the chain of conclusions, it could
only be the assumption that was wrong. Therefore, there must be an infinite
number of primes!\\

\hypertarget{euclid}{}
\index{Euclid's proof by contradiction}\index{proof by contradiction}
% \paragraph*{Euclid's proof by contradiction}
% goes as follows:
\begin{proof}{according to Euclid (proof by contradiction)\\}
{\bf Assumption:} \quad There is a {\em finite} number of primes. \\*[4pt] {\bf
Conclusion:} \quad Then these can be listed $p_1 < p_2 < p_3 < \dots < p_n$,
where $n$ is the (finite) number of prime numbers. $p_n$ is therefore the
largest prime. Euclid now looks at the number $a = p_1 \cdot p_2 \cdots p_n +1$.
This number cannot be a prime number because it is not included in our list of
primes. It must therefore be divisible by a prime, i.e. there is a natural
number $i$ between $1$ and $n$, such that $p_i$ divides the number $a$. Of
course, $p_i$ also divides the product $a-1 = p_1 \cdot p_2 \cdots p_n$, because
$p_i$ is a factor of $a-1$. Since $ p_i $ divides the numbers $ a $ and $ a-1 $,
it also divides the difference of these numbers. Thus: $p_i$ divides  $a - (a-1)
= 1$. $p_i$ must therefore divide $1$, which is impossible. \\*[4pt] {\bf
Contradiction:} \quad Our assumption was false.

@@

The largest prime numbers known today have several millions digits, 
which is too big for us to imagine. The number of elementary particles in
the universe is estimated to be ``only'' a $80$-digit number 
\hyperlink{grosord}{(See: overview under \ref{s:grosord}
of various orders of magnitude / dimensions)}.

% --------------------------------------------------------------------------
\hypertarget{RecordPrimes}{}
\subsection{The 20+ largest known primes (as of January 2016)}  % Eyecatcher_neue-Mersenne
\label{RecordPrimes}
\index{prime number!records}

The following table contains the biggest, currently known primes and
a description of its particular number type\footnote{%
An up-to-date version can be found in the internet
at   \url{http://primes.utm.edu/largest.html},
at   \url{http://primes.utm.edu/mersenne/index.html}, and
at   \url{http://www.mersenne.org/primes/}.
}:

\index{Mersenne!number!generalized}
\index{Fermat!number!generalized}

\begin{table} % don't use [h] (error if table has >24 lines)
\begin{center}
\begin{tabular}{|c|cccc|}
\hline \rule{0pt}{10pt}    % Eyecatcher_neue-Mersenne
	& {\bf Definition} & {\bf Decimal Digits} & {\bf When} & {\bf Description} \\
\hline \rule{0pt}{15pt}
	1  & $2^{74,207,281}-1$ & 22,338,618 & 2016 & Mersenne, 49th known \\
	2  & $2^{57,885,161}-1$ & 17,425,170 & 2013 & Mersenne, 48th known \\
	3  & $2^{43,112,609}-1$ & 12,978,189 & 2008 & Mersenne, 47th known \\
	4  & $2^{42,643,801}-1$ & 12,837,064 & 2009 & Mersenne, 46th known \\
	5  & $2^{37,156,667}-1$ & 11,185,272 & 2008 & Mersenne, 45th known \\
	6  & $2^{32,582,657}-1$ &  9,808,358 & 2006 & Mersenne, M-44 \\
	7  & $2^{30,402,457}-1$ &  9,152,052 & 2005 & Mersenne, M-43 \\
	8  & $2^{25,964,951}-1$ &  7,816,230 & 2005 & Mersenne, M-42 \\
	9  & $2^{24,036,583}-1$ &  7,235,733 & 2004 & Mersenne, M-41 \\
	10 & $2^{20,996,011}-1$ &  6,320,430 & 2003 & Mersenne, M-40 \\
	11 & $2^{13,466,917}-1$ &  4,053,946 & 2001 & Mersenne, M-39 \\
	12 & $19,249 \cdot 2^{13,018,586}+1$ & 3,918,990 & 2007 & Generalized Mersenne\footnotemark \\
	13 & $27,653 \cdot 2^{9,167,433}+1$ & 2,759,677 & 2005 & Generalized Mersenne \\
	14 & $28,433 \cdot 2^{7,830,457}+1$ & 2,357,207 & 2004 & Generalized Mersenne \\
	15  & $2^{ 6,972,593}-1$ & 2,098,960 & 1999 & Mersenne, M-38 \\
	16  & $5,359 \cdot 2^{5,054,502}+1$ & 1,521,561 & 2003 & Generalized Mersenne \\
	17  & $4,847 \cdot 2^{3,321,063}+1$ & 999,744 & 2005 & Generalized Mersenne \\
	18  & $3 \cdot 2^{3,136,255}-1$ & 944,108 & 2007 & Generalized Mersenne \\
	19  & $2^{ 3,021,377}-1$ &   909,526 & 1998 & Mersenne, M-37 \\
	20  & $2^{ 2,976,221}-1$ &   895,932 & 1997 & Mersenne, M-36 \\
	21  & $222,361 \cdot 2^{2,854,840}+1$ & 859,398 & 2006 & Generalized Mersenne \\
	22 & $1,372,930^{131,072}+1$ &   804,474 & 2003 & Generalized Fermat\footnotemark \\
	23  & $1,361,244^{131,072}+1$ &   803,988 & 2004 & Generalized Fermat \\
	24  & $1,176,694^{131,072}+1$ &   795,695 & 2003 & Generalized Fermat \\
	25  & $342,673 \cdot 2^{2,639,439}-1$ & 794,556 & 2007 & Generalized Mersenne \\

\hline
\end{tabular}
\caption{The 20+ largest known primes and its particular number types
         (as of January 2016)}    % Eyecatcher_neue-Mersenne
\label{L_n_Largest_Known-Primes}
\end{center}
\end{table} 
\addtocounter{footnote}{-1}
\footnotetext{\index{number!Sierpinski}\index{Seventeen or Bust SoB}%
This number was found within the distributed computing project
``Seventeen or Bust'' (SoB) (\url{http://www.seventeenorbust.com})
% {\href{http://www.mersenne.org} {\tt http://www.mersenne.org}}
at March 26, 2007. While the well known \hyperlink{GIMPS-project}{GIMPS project} (chapter~\ref{zahlentyp_mersenne}) searches for bigger and bigger of the infinitely many primes, there is a chance, that the SoB project could have been completed its task sometime.

@@

When the project started in 2002 there have been 17 possible candidates $< 78557$ (this is the reason for the project's name ``Seventeen or Bust''). It is sufficient to find one single counter-example, to exclude a candidate k, which means to find a single $n>=1$, where $N = k \cdot 2^{n}+1$ is prime. So it is only a byproduct of this task that this also generates new monster primes.
} 
\addtocounter{footnote}{1}
\footnotetext{%
Generalized Fermat number: $ 1,372,930^{131,072} + 1 = 1,372,930^{(2^{17})}+1 $
\index{Fermat!number!generalized}  } 
%be_2005: Erzwingen, dass die Abb. noch in diesem Kapitel !

The largest known prime is a Mersenne prime, found by the
\hyperlink{GIMPS-project}{GIMPS project}(chapter~\ref{zahlentyp_mersenne}).

Within the largest known primes there are also numbers of the type
\hyperlink{generalizedMersennenumbers}{generalized Mersenne number}
(chapter~\ref{generalized-mersenne-no1})
and 
\hyperlink{generalizedFermatprimes}{generalized Fermat numbers}
(chapter~\ref{generalized-fermat}).

@@

\vskip + 5pt
\hypertarget{Mer-nums-not-always-prim}{}
Unfortunately this theorem only applies in one direction (the inverse 
statement does not apply, no equivalence): that means that there exist 
prime exponent for which the Mersenne number is {\bf not} prime (see the 
above example $2^{11}-1, $ where $11$ is prime, but $2^{11}-1$ not).

Mersenne claimed that $2^{67}-1$ is a prime number. There is also a mathematical
history behind this claim: it first took over 200 years before 
\index{Lucas, Edouard} Edouard Lucas (1842-1891) proved that this number 
is composite.
However, he argued indirectly and did not name any of the factors. In 1903,
Cole\index{Cole, Frank Nelson}\footnote{%
Frank Nelson Cole, American mathematician, Sep. 20, 1861 $-$ May 26, 1926.}
showed which factors make up this composite number: 
$$ 2^{67} -1
=147, 573, 952, 589, 676, 412, 927 = 193, 707, 721 \cdot 761, 838, 257, 287. $$
He admitted to having worked 20 years on the factorization 
\index{factorization} (expression as a product of prime factors)\footnote{%
  Using CT1\index{CrypTool 1} you can factorize numbers in the 
  following way: menu {\bf Indiv. Procedures \textbackslash{} RSA Cryptosystem 
  \textbackslash{} Factorization of a Number}. \\
  CT1 can factorize with the quadratic sieve (QS) on a single PC in a reasonable
  time numbers no longer than 250 bit.
  Numbers bigger than 1024 bits are currently not accepted by CT1. \\
  With CT2 you can use both the QS and the GNFS (general number field sieve),
  and CT2 can distribute the calculations on several computers.\\
  The current factorization records are listed in chapter \ref{nt:NoteFactorization}.
  \index{factorization!factoring records}
}
of this 21-digit decimal number!

>>>>

% $Id$
% ..............................................................................
%             CHAPTER 3:  P R I M E S
%
% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

\begin{bibunit}[babalpha] %% alpha: Chapter bibliography shows authors abbreviation

\newpage
\hypertarget{Chapter_Primes}{}

\chapter{Prime Numbers}
\label{Chapter_Primes}
(\hyperlink{author_Bernhard-Esslinger}{Bernhard Esslinger}, May 1999;
 Updates: Nov 2000, Dec 2001, Jun 2003, May 2005, Mar 2006, Jun 2007,
          Jan 2010, Aug 2013, Jul 2016)

@@

\noindent Prime numbers are particularly important for three reasons:
\begin{itemize}
  \item In number theory, they are considered to be the basic components of
natural numbers, upon which numerous brilliant mathematical ideas are based.
  \item They are of extreme practical importance in modern
\index{cryptography!modern} cryptography (public key \index{cryptography!public key} cryptography). The most common public key procedure, invented at the end of
the 1970's, is \index{RSA} RSA encryption. Only using (large) prime numbers for
particular parameters can you guarantee that an algorithm is secure, both for
the RSA procedure and for even more modern procedures (e.g. elliptic curves).
  \item The search for the largest known prime numbers does not have any
practical usage known to date, but requires the best computers, is an excellent
benchmark (possibility for determining the performance\index{performance} of computers) and leads
to new calculation methods on many computers (see also: \url{http://www.mersenne.org/prime.htm}).
\end{itemize}

Many people have been fascinated by prime numbers over the past two millennia.
Ambition to make new discoveries about prime numbers has often resulted in
brilliant ideas and conclusions. The following section provides an easily
comprehensible introduction to the basics of prime numbers. We will also
explain what is known about the distribution (density, number of prime numbers
in particular intervals) of prime numbers and how prime number tests work.

@@

\begin{figure}[!hb]
\begin{center}
\includegraphics[scale=0.3]{figures/Numberrectangle-with-colored-primes_Irisprime.jpg}
\caption[Primes within the first 390 integers -- marked with color]
        {Primes within the first 390 integers -- marked with color\protect\footnotemark}
\label{Primes-in-a-390-integer-rechtangle-figure}
\end{center}
\end{figure}
\footnotetext{%
   Graphics from 
   \url{http://mathforum.org/mathimages/index.php/Image:Irisprime.jpg}, 30*13 rectangle
   }

\begin{figure}[!hb]
\begin{center}
\includegraphics[scale=0.41]{figures/Ulam_spiral_via-CT2-within-1-999.png}
\caption[Primes within the first 999 integers -- as Ulam spiral]
        {Primes within the first 999 integers -- as Ulam spiral\protect\footnotemark}
\label{Primes-in-a-999-integer-ulam-spiral-figure}
\end{center}
\end{figure}
\footnotetext{%
   Graphics from CT2, menu Crypto Tutorials, World of Primes,
   Distribution of primes, Ulam's spiral; 32*32 points.\index{CT2}
}
%\newpage
\begin{figure}[!hb]
\begin{center}
\includegraphics{figures/Ulam_spiral_Wikipedia.png}
\caption[Primes within the first 4000 integers -- as Ulam spiral]
        {Primes within the first 4000 integers -- as Ulam spiral\protect\footnotemark}
\label{Primes-in-a-4000-integer-ulam-spiral-figure}
\end{center}
\end{figure}
\footnotetext{%
   Graphics from 
   \url{http://mathforum.org/mathimages/index.php/Image:Ulam_spiral.png}, 200*200 Ulam spiral
   }

@@

% --------------------------------------------------------------------------
\section{How many prime numbers are there?}
\label{Section_pr_How-many-primes}

For the natural numbers, the primes can be compared to elements in chemistry or
the elementary particles in physics (see \cite[p. 22]{Blum1999}).

Although there are only $92$ natural chemical elements, the number of prime
numbers is unlimited. Even the Greek, \index{Euclid} Euclid%
\footnote{Euclid,
a Greek mathematician of 4th and 3rd century B.C. He worked at the
Egyptian academy of Alexandria and wrote ``The Elements'', the most well 
known systematically textbook of the Greek mathematics.} 
knew this in the third century B.C.

\begin{theorem}[Euclid\footnote{The common usage of the term does not denote Euclid as the inventor of the theorem rather;
the true inventor is merely not as prominent. The theorem has already been distinguished
and proven in Euclid's Elements (Book IX, theorem 20). The phraseology is remarkable due to 
the fact that the word infinite is not used. The text reads as followed
$$
O\acute{\iota}~\pi\varrho\tilde{\omega}\tau o \iota~\grave{\alpha}\varrho\iota\vartheta\mu o\grave{\iota}~
\pi\lambda\varepsilon\acute{\iota}o \upsilon\varsigma~\varepsilon\grave{\iota}\sigma\grave{\iota}~
\pi\alpha\nu\tau\grave{o}\varsigma~\tau o \tilde{\upsilon}~
\pi\varrho o \tau\varepsilon\vartheta\acute{\varepsilon}\nu\tau o \varsigma~
\pi\lambda\acute{\eta}\vartheta\ o \upsilon\varsigma~
\pi\varrho\acute{\omega}\tau\omega\nu~
\grave{\alpha}\varrho\iota\vartheta\mu\tilde{\omega}\nu,
$$
the English translation of which is: the prime numbers are more than 
any previously existing amount of prime numbers.
}]\label{thm-pz-euklid} % Ende der Fuﬂnote
% Fussnote VOR "]" in [Euclid], damit kein Leerraum vor der Fussnotennummer.
% Vorher stand da: ...(Euclid). BLANK Fussnote und das Blank stoerte.
% Nun steht die Fussnote direkt hinter "Euclid" und vor der ")".
% Eigentlich h‰tte ich sie gerne direkt hinter "(Euclid)", noch vor dem 
% automatisch gesetzten Punkt. 
The sequence of prime numbers does not discontinue.
Therefore, the quantity of prime numbers is infinite.
\end{theorem}

His proof that there is an infinite number of
primes is still considered to be a brilliant mathematical consideration and
conclusion today (\textbf{proof by contradiction} \index{proof by contradiction}). 
He assumed that there is only a finite number of primes and therefore a 
largest prime number. Based on this assumption, he drew logical conclusions
until he obtained an obvious contradiction. This meant that something must
be wrong. As there were no mistakes in the chain of conclusions, it could
only be the assumption that was wrong. Therefore, there must be an infinite
number of primes!\\

\hypertarget{ht_euclid}{}
\index{Euclid's proof by contradiction}\index{proof by contradiction}
% \paragraph*{Euclid's proof by contradiction}
% goes as follows:
\begin{proof}{according to Euclid (proof by contradiction)\\}
{\bf Assumption:} \quad There is a {\em finite} number of primes. \\*[4pt] {\bf
Conclusion:} \quad Then these can be listed $p_1 < p_2 < p_3 < \dots < p_n$,
where $n$ is the (finite) number of prime numbers. $p_n$ is therefore the
largest prime. Euclid now looks at the number $a = p_1 \cdot p_2 \cdots p_n +1$.
This number cannot be a prime number because it is not included in our list of
primes. It must therefore be divisible by a prime, i.e. there is a natural
number $i$ between $1$ and $n$, such that $p_i$ divides the number $a$. Of
course, $p_i$ also divides the product $a-1 = p_1 \cdot p_2 \cdots p_n$, because
$p_i$ is a factor of $a-1$. Since $ p_i $ divides the numbers $ a $ and $ a-1 $,
it also divides the difference of these numbers. Thus: $p_i$ divides  $a - (a-1)
= 1$. $p_i$ must therefore divide $1$, which is impossible. \\*[4pt] {\bf
Contradiction:} \quad Our assumption was false.

@@

The largest prime numbers known today have several millions digits, 
which is too big for us to imagine. The number of elementary particles in
the universe is estimated to be ``only'' a $80$-digit decimal number 
\hyperlink{grosord}{(see the overview under \ref{s:grosord}
of various orders of magnitude / dimensions)}.

% --------------------------------------------------------------------------
\hypertarget{RecordPrimes}{}
\subsection{The 30+ largest known primes (as of June 2016)}  % Eyecatcher_neue-Mersenne
\label{RecordPrimes}
\index{prime number!records}

The following table contains the biggest, currently known primes and
a description of its particular number type\footnote{%
An up-to-date version can be found in the internet
at   \url{https://primes.utm.edu/primes/search.php?Number=1000},
at   \url{http://primes.utm.edu/mersenne/index.html}, and
at   \url{http://www.mersenne.org/primes/}.
}:

\index{Mersenne!number!generalized}
\index{Fermat!number!generalized} 
\begin{table}[ht] % don't use [h] (error if table has >24 lines)
\begin{center}
\begin{tabular}{|c|cccc|}
\hline \rule{0pt}{10pt}    % Eyecatcher_neue-Mersenne
	& {\bf Definition} & {\bf Decimal Digits} & {\bf When} & {\bf Description} \\
\hline \rule{0pt}{15pt}
	1  & $2^{74,207,281}-1$ & 22,338,618 & 2016 & Mersenne, 49th known \\
	2  & $2^{57,885,161}-1$ & 17,425,170 & 2013 & Mersenne, 48th known \\
	3  & $2^{43,112,609}-1$ & 12,978,189 & 2008 & Mersenne, 47th known \\
	4  & $2^{42,643,801}-1$ & 12,837,064 & 2009 & Mersenne, 46th known \\
	5  & $2^{37,156,667}-1$ & 11,185,272 & 2008 & Mersenne, 45th known \\
	6  & $2^{32,582,657}-1$ &  9,808,358 & 2006 & Mersenne, M-44 \\
	7  & $2^{30,402,457}-1$ &  9,152,052 & 2005 & Mersenne, M-43 \\
	8  & $2^{25,964,951}-1$ &  7,816,230 & 2005 & Mersenne, M-42 \\
	9  & $2^{24,036,583}-1$ &  7,235,733 & 2004 & Mersenne, M-41 \\
	10 & $2^{20,996,011}-1$ &  6,320,430 & 2003 & Mersenne, M-40 \\
	11 & $2^{13,466,917}-1$ &  4,053,946 & 2001 & Mersenne, M-39 \\
	12 & $19,249 \cdot 2^{13,018,586}+1$ & 3,918,990 & 2007 & Generalized Mersenne\footnotemark \\%%%

	13 & $3 \cdot 2^{11,895,718}-1$ & 3,580,969 & 2015 & Generalized Mersenne \\
	14 & $3 \cdot 2^{11,731,850}-1$ & 3,531,640 & 2015 & Generalized Mersenne \\
	15 & $3 \cdot 2^{11,484,018}-1$ & 3,457,035 & 2014 & Generalized Mersenne \\
	16 & $3 \cdot 2^{10,829,346}+1$ & 3,259,959 & 2014 & Generalized Mersenne \\
	17 & $ 475,856^{524,288}+1$ &   2,976,633 & 2012 & Generalized Fermat \\
	18 & $ 356,926^{524,288}+1$ &   2,911,151 & 2012 & Generalized Fermat \\
	19 & $ 341,112^{524,288}+1$ &   2,900,832 & 2012 & Generalized Fermat \\

	20 & $27,653 \cdot 2^{9,167,433}+1$ & 2,759,677 & 2005 & Generalized Mersenne \\%%%

	21 & $90,527 \cdot 2^{9,162,167}+1$ & 2,758,093 & 2010 & Generalized Mersenne \\
	22 & $2,038 \cdot 366^{1,028,507}-1$ & 2,636,562 & 2016 & Generalized Fermat \\
	23 & $75,898^{524,288}+1$ & 2,558,647 & 2011 & Generalized Fermat \\

	24 & $28,433 \cdot 2^{7,830,457}+1$ & 2,357,207 & 2004 & Generalized Mersenne \\%%%

	25 & $502,573 \cdot 2^{7,181,987}-1$ & 2,162,000 & 2014 & Generalized Mersenne \\
	26 & $402,539 \cdot 2^{7,173,024}-1$ & 2,159,301 & 2014 & Generalized Mersenne \\
	27 & $161,041 \cdot 2^{7,107,964}+1$ & 2,139,716 & 2015 & Generalized Mersenne \\
	28 & $3 \cdot 2^{7,033,641}+1$ & 2,117,338 & 2011 & Generalized Mersenne \\
	29 & $33,661 \cdot 2^{7,031,232}+1$ & 2,116,617 & 2007 & Generalized Mersenne \\

	30  & $2^{ 6,972,593}-1$ & 2,098,960 & 1999 & Mersenne, M-38 \\
        ... &  &  &  & \\
%	16  & $5,359 \cdot 2^{5,054,502}+1$ & 1,521,561 & 2003 & Generalized Mersenne \\
%	17  & $4,847 \cdot 2^{3,321,063}+1$ & 999,744 & 2005 & Generalized Mersenne \\
%	18  & $3 \cdot 2^{3,136,255}-1$ & 944,108 & 2007 & Generalized Mersenne \\
%	19  & $2^{ 3,021,377}-1$ &   909,526 & 1998 & Mersenne, M-37 \\
%	20  & $2^{ 2,976,221}-1$ &   895,932 & 1997 & Mersenne, M-36 \\
%	21  & $222,361 \cdot 2^{2,854,840}+1$ & 859,398 & 2006 & Generalized Mersenne \\
	328 & $1,372,930^{131,072}+1$ &   804,474 & 2003 & Generalized Fermat\footnotemark \\
%	23  & $1,361,244^{131,072}+1$ &   803,988 & 2004 & Generalized Fermat \\
%	24  & $1,176,694^{131,072}+1$ &   795,695 & 2003 & Generalized Fermat \\
        ... &  &  &  & \\
	342  & $342,673 \cdot 2^{2,639,439}-1$ & 794,556 & 2007 & Generalized Mersenne \\

\hline
\end{tabular}
\caption{The 30+ largest known primes and its particular number types
         (as of June 2016)}    % Eyecatcher_neue-Mersenne
\label{L_n_Largest_Known-Primes}
\end{center}
\end{table} 
\addtocounter{footnote}{-1}
\footnotetext{\index{number!Sierpinski}\index{Seventeen or Bust SoB}%
This number was found within the distributed computing project
``Seventeen or Bust'' (SoB)  (\url{https://en.wikipedia.org/wiki/Seventeen_or_Bust})
at March 26, 2007. While the well known \hyperlink{GIMPS-project}{GIMPS project} (chapter~\ref{zahlentyp_mersenne}) searches for bigger and bigger of the infinitely many primes, there is a chance, that the SoB project could have been completed its task sometime.

@@

When the project started in 2002 there have been 17 possible candidates $< 78557$ (this is the reason for the project's name ``Seventeen or Bust''). It is sufficient to find one single counter-example, to exclude a candidate k, which means to find a single $n>=1$, where $N = k \cdot 2^{n}+1$ is prime. So it is only a byproduct of this task that this also generates new monster primes.

As of about April 19, 2016, the main SoB server is down and the future of the project unknown.
} 
\addtocounter{footnote}{1}
\footnotetext{%
Generalized Fermat number: $ 1,372,930^{131,072} + 1 = 1,372,930^{(2^{17})}+1 $
\index{Fermat!number!generalized}  }

The development over time is shown in figure \ref{Figure_Caldwell_Largest-Known-Prime-From1975}.
It starts with M21701 found in 1978. Note the vertical scale.
\begin{figure}[ht]
\begin{center}
\includegraphics[scale=0.7]{figures/Caldwell_From1975-2016-en.png}
\caption[Number of digits of largest known prime by year since 1975]
        {Number of digits of largest known prime by year (as of July 2016)\footnotemark}
\label{Figure_Caldwell_Largest-Known-Prime-From1975}
\end{center}
\end{figure}
\index{Caldwell, Chris}
\footnotetext{
  Source: Chris Caldwell, \url{http://primes.utm.edu/notes/by_year.html}
}

The larges currently known prime is a Mersenne prime, found by the
\hyperlink{GIMPS-project}{GIMPS project} (chapter~\ref{zahlentyp_mersenne}).
Within the largest known primes there are also numbers of the type
\hyperlink{generalizedMersennenumbers}{generalized Mersenne number}
(chapter~\ref{generalized-mersenne-no1})
and 
\hyperlink{generalizedFermatprimes}{generalized Fermat numbers}
(chapter~\ref{generalized-fermat}).

@@

\vskip + 5pt
\hypertarget{Mer-nums-not-always-prim}{}
Unfortunately this theorem only applies in one direction (the inverse 
statement does not apply, no equivalence): That means that there exist 
prime exponent for which the Mersenne number is {\bf not} prime (see the 
above example $2^{11}-1, $ where $11$ is prime, but $2^{11}-1$ not).

Mersenne claimed that $2^{67}-1$ is a prime number. There is also a mathematical
history behind this claim: It first took over 200 years before 
\index{Lucas, Edouard} Edouard Lucas (1842-1891) proved that this number 
is composite.
However, he argued indirectly and did not name any of the factors. In 1903,
Cole\index{Cole, Frank Nelson}\footnote{%
Frank Nelson Cole, American mathematician, Sep. 20, 1861 $-$ May 26, 1926.\\
}
showed which factors make up this composite number: 
$$ 2^{67} -1
=147, 573, 952, 589, 676, 412, 927 = 193, 707, 721 \cdot 761, 838, 257, 287. $$
He admitted to having worked 20 years on the {\bf factorization}
\index{factorization} (dissection as a product of prime factors)\footnote{%
  Using CT1\index{CT1} you can factorize numbers in the 
  following way: menu {\bf Indiv. Procedures \textbackslash{} RSA Cryptosystem 
  \textbackslash{} Factorization of a Number}.\\
  CT1 can factorize with the quadratic sieve (QS) on a single PC in a reasonable
  time numbers no longer than 250 bit. Numbers bigger than 1024 bits are anyway
  currently not accepted by CT1.\\

  \noindent CT2\index{CT2} has a GeneralFactorizer component
  based on YAFU\index{YAFU}. This is faster than the implemented functions in CT1.
  So CT2 offers the following factoring methods:
  \begin{compactitem}
   \item[-] brute-force with small primes
   \item[-] Fermat
   \item[-] Shanks square forms factorization (squfof)
   \item[-] Pollard rho
   \item[-] Pollard p-1
   \item[-] Williams p+1
   \item[-] Lenstra elliptic curve method (ECM)
   \item[-] self-initializing quadratic sieve (SIQS)
   \item[-] multiple polynomial quadratic sieve (MPQS)
   \item[-] special number field sieve (SNFS)
   \item[-] general number field sieve (GNFS).         
  \end{compactitem}

  \noindent CT2 started to experiment with a general infrastructure for distributed computing
  called CrypCloud\index{CrypCloud} (both peer-to-peer and centralized). So in the future,
  CT2 will be able to distribute the calculations on many computers. What could be achieved
  after the components are made ready for parallelization showed a cluster for
  distributed cryptanalysis of DES and AES: Status on March 21st, 2016 is,
  that an AES brute-force attack (distributed keysearching) worked on
  50 i5 PCs, each with 4 virtual CPU cores. These 200 virtual ``worker threads'' 
  achieved to test about 350 million AES keys/sec. The ``cloud'' processed a total
  amount of about 20 GB/sec of data. CrypCloud is a peer-to-peer-based
  cloud system that enables CT2 users to voluntarily join distributed computing jobs.

  The current factorization records are listed in chapter \ref{nt:NoteFactorization}.
  \index{factorization!factoring records}\\
}
of this 21-digit decimal number!
